---
title: "Practical 4"
format: html
editor: visual
---

## Statistical Modelling: Practical 4

In this practical we will discuss model selection. The data are provided in the file `Stackloss.txt` (available on KEATS).

Below we load the data, inspect the first few rows, and attach the data frame.

```{r}
# Load data
Data <- read.table("Stackloss.txt", header = TRUE)
head(Data)

# Attach the data frame for easy variable access
attach(Data)
```

### 1. Partial F-tests

A partial F test tells us whether a larger model is explaining significantly more variance than the smaller model accounting for the additional degrees of freedom in the larger model.

**Steps for Partial** $F$**-test**

1.  Fit two *nested* models:
    -   A full model (e.g. $Y \sim X_1 + X_2 + X_3$)
    -   A reduced model (e.g., $Y \sim X_1$)
2.  Compute the partial $F$-statistic using ANOVA:

$$F = \frac{(SSR_{\text{reduced}} - SSR_{\text{full}}) / (df_{\text{reduced}} - df_{\text{full}})}{SSE_{\text{full}} / df_{\text{full residual}}} $$

-   $SSR$: Regression sum of squares
-   $SSE$: Error sum of squares
-   $df$: Degrees of freedom

3.  Evaluate the $p$-value:
    -   If $p < 0.05$, the additional predictors significantly improve the model fit.
    -   If $p \geq 0.05$, the simpler model is sufficient.

We begin by fitting the full model using all three predictors.

```{r}
# Full model with all predictors
ModelFull <- lm(Stkloss ~ Air + Temp + Acid)
summary(ModelFull)
```

Next, we fit a reduced model using only `Temp` as the predictor.

```{r}
# Reduced model with only Temp as predictor
ModelRed <- lm(Stkloss ~ Temp)
```

We then carry out the partial $F$-test to compare the two models.

```{r}
# Partial F-test between the reduced and full model
anova(ModelRed, ModelFull)
```

Discussion: Based on the partial $F$-test output, discuss whether the additional predictors (`Air` and `Acid`) significantly improve the model over using `Temp` alone.

### 2. Comparison of Non-Nested Models

Since the partial $F$-test only works for nested models, we now compare non-nested models. Consider the following model:

```{r}
# Model with predictors Air and Acid (excluding Temp)
Model2 <- lm(Stkloss ~ Air + Acid)
summary(Model2)
```

Discussion: Review the summaries of the models. Which model appears better based on the adjusted $R^2$ values?

**Multiple R-squared vs. Adjusted R-squared**

Multiple $R^2$ measures the proportion of variability in the response variable explained by the regression model: $R^2 = 1 - \frac{SSE}{SST}$ where:

-   $SSE$: Sum of Squared Errors
-   $SST$: Total Sum of Squares

Adjusted R-squared adjusts (R\^2) for the number of predictors, penalizing more complex models: $R^2_{\text{adj}} = 1 - \left[ \frac{SSE / (n - p - 1)}{SST / (n - 1)} \right]$ where:

-   $n$: Sample size
-   $p$: Number of predictors

Comparison:

-   Multiple $R^2$ always increases or stays the same when adding predictors, potentially overestimating model quality

-   Adjusted $R^2$ can decrease if extra predictors don't sufficiently improve the model, providing a more balanced measure for model comparison

### 3. Comparison Using AIC

Another way to compare models is by using the Akaike Information Criterion (AIC). AIC is a measure for model selection that balances model fit and complexity. It is defined as:
$\text{AIC} = 2k - 2\ln(\hat{L})$ where:

-   $k$: Number of estimated parameters in the model

-   $\hat{L}$: Maximum value of the likelihood function for the model, that is the likelihood given by $Y \sim N(X\beta, \sigma^2 I)$ where the parameters $\beta$ and $\sigma^2$ for each model are estimated using the Maximum Likelihood Estimates.

We calculate the AIC for the three models.

```{r}
# Compute AIC for each model
AIC(ModelRed)
AIC(Model2)
AIC(ModelFull)
```

Discussion: Which model is preferred based on the AIC values?\
*Optional:* *You might also try listing all possible models using the three predictors and choose the best one based on AIC.*

### 4. Predictive $R^2$

Predictive $R^2$ measures how well a regression model predicts new observations. It is defined as:$R^2_{\text{pred}} = 1 - \frac{\text{PRESS}}{\text{SST}}$.
The Predictive Residual Sum of Squares (PRESS) is calculated as:

$\text{PRESS} = \sum_{i=1}^{n} \big(\frac{e_i}{1-h_{ii}}\big)^2$

where

-   $e_i$: The residual for observation $i$

-   $h_{ii}$: the leverage value taken from the diagonal of the hat matrix for observation $i$

-   $SST$: The total sum of squares given by $\sum_{i=1}^{n} (y_i - \bar{y})^2$

Interpretation:

-   Higher predictive $R^2$ indicates better model performance

-   Unlike regular $R^2$, predictive $R^2$ assess the model's predictive ability on unseen data, assessing predictive performance.

Predictive $R^2$ is not provided directly by R. We define a function to compute it.

```{r}
pred.R2 <- function(model) {
  # Obtain the design matrix X of the model.
  X <- model.matrix(model)
  # Compute the projection matrix H.
  H <- X %*% solve(t(X) %*% X) %*% t(X)
  # Extract the diagonal elements of H.
  h <- diag(H)
  # Compute the predictive residuals (PRESS).
  PRESS <- sum((model$residuals / (1 - h))^2)
  # Compute the total sum of squares.
  SST <- sum((model$model[,1] - mean(model$model[,1]))^2)
  # Calculate the predictive R².
  Rp <- 1 - PRESS / SST
  return(Rp)
}
```

Now, we calculate the predictive $R^2$ for each model.

```{r}
pred.R2(ModelFull)
pred.R2(ModelRed)
pred.R2(Model2)
```

Discussion: Based on the predictive $R^2$ values, which model provides the best predictive performance?

# 5. Step-wise Search Based on AIC

Finally, we perform a step-wise model search using AIC to potentially find a better model.

```{r}
# Perform step-wise search starting from the full model
step(ModelFull)
```

Alternatively, you can save the best model and review its details.

```{r}
# Save the best model from step-wise search
Model.best <- step(ModelFull)
summary(Model.best)
plot(Model.best)
```

Discussion: Review the results from the step-wise search and compare the best model with the models you have already fitted.
