---
title: "DistributionResults"
format: html
editor: visual
---

## Distribution Results

As in the tutorial, let's simulate four sets of observations from three standard normal distributions.

```{r}
# First let's set the parameters mu and sigma
mu = 0
sigma = 1
n = 4

# Now we get our three sets of four observations
z1_obs = rnorm(n, mean=mu, sd = sigma)
z2_obs = rnorm(n, mean=mu, sd = sigma)
z3_obs = rnorm(n, mean=mu, sd = sigma)

# Let's collect our observations in a matrix
obs_mat = matrix(c(z1_obs, z2_obs, z3_obs), 4, 3)
obs_mat
```

Now we can square the values and sum them together to get four values sampled from a sum of 3 squared normals

```{r}
sq_obs_mat = obs_mat**2
sum_squares = rowSums(sq_obs_mat)
sum_squares
```

The mathematics suggests that these four observations come from a $\chi^2_3$ distribution. We can validate this by repeating the procedure but for 1000 observations rather than 4.

```{r}
n=1000
z1_obs = rnorm(n, mean=mu, sd = sigma)
z2_obs = rnorm(n, mean=mu, sd = sigma)
z3_obs = rnorm(n, mean=mu, sd = sigma)

obs_mat = matrix(c(z1_obs, z2_obs, z3_obs), n, 3)
sq_obs_mat = obs_mat**2
sum_squares = rowSums(sq_obs_mat)

# Let's plot a histogram of the results
hist(sum_squares, breaks = 70, xlim = c(0, 15), main='Sample from Normal Dists')

# We can compare this to a collection of obsevations taken directly from the chi-squared distribution
chisq_sample = rchisq(1000000, 3)
hist(chisq_sample, breaks = 150, xlim = c(0, 15), main='Sample from Chi-Squared')
```

We can also compare Kernel Density Estimates of these two sets of variables. Kernel density estimates plot the empirical density of some observed values - conceptually you can think of this as like smoothing out a histogram.

```{r}
plot(density(chisq_sample), col='blue')
lines(density(sum_squares), col='red')
legend('topright', legend = c('Chi Squared', 'Normals'), col = c('blue', 'red'), lty = 1)
```

See how the two curves are very similar, suggesting similar distributions.

Similar sampling procedures can be used to empirically demonstrate that if $U \sim \chi^2_m$ and $V\sim \chi^2_n$, then

-   $$\frac{U/m}{V/n} \sim F_{m,n}$$

-   $$\frac{Z_i}{\sqrt{V/n}} \sim t_n$$

-   

### Exercise

As an exercise, see if you can use the `rchisq(n, df)`, `rf(n, df_1, df_2)` and `rt(n, df)` functions to generate random samples from a chi-squared, F, and student's-t distributions respectively alongside kernel density estimates to demonstrate these mathematical results empirically.
